---
title: "Visualization Draft"
format:
  pdf: 
    fig-width: 10
    fig-height: 8
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
editor: source
execute: 
  message: false
  warning: false
---

# 1. Set up

## Packages

```{r}
#| label: packages
#| message: false

# general use: data wrangling and plotting
library(tidyverse)

# clean column names
library(janitor)

# file organization
library(here)

# working with dates and times
library(lubridate)
```

## Data

```{r}
#| label: data
#| message: false

# weather data from NOAA weather station
weather <- read_csv(here("data", "NOAA-weather-data.csv"))

# water quality survey metadata
metadata <- read_csv(here("data", "NCOS_YSI_Water_Quality_Monitoring_0.csv"))

# water quality parameter measurements
water_parameters <- read_csv(here("data", "YSI_Data_Begin_1.csv"))
```

# 2. Clean data

## Cleaning weather data

```{r}
#| label: cleaning-weather-data

# creating new object from weather
weather_clean <- weather |> 
  # cleaning column names
  clean_names() |> 
  # making sure date is in date format
  mutate(date = mdy(date)) |> 
  # selecting columns of interest
  select(date, prcp, tmax, tmin) |> 
  # extracting month and year from date column
  mutate(
    month = month(date),
    year = year(date)
  ) |> 
  # assigning water year
  mutate(
    wy = case_when(
      # when month is October or later, assign value of year + 1
      month >= 10 ~ year + 1,
      # when month is September or earlier, assign value of existing year
      TRUE ~ year
    )
  )
```

## Cleaning metadata

```{r}
#| label: cleaning-metadata

# creating new object from metadata
metadata_clean <- metadata |> 
  # cleaning column names
  clean_names() |> 
  # excluding unnecessary columns
  select(-any_of(c("object_id", "creator", "editor"))) |> 
  # making sure monitoring_date is in date-time format
  mutate(monitoring_date = mdy_hms(monitoring_date)) |> 
  # renaming monitoring datetime column to reflect values
  rename(monitoring_datetime = monitoring_date) |> 
  # creating new columns for month and year of observations
  mutate(
    month = month(monitoring_datetime),
    year = year(monitoring_datetime)
  ) |> 
  # assigning water year
  mutate(
    wy = case_when(
      # when month is October or later, assign value of year + 1
      month >= 10 ~ year + 1,
      # when month is September or earlier, assign value of existing year
      TRUE ~ year
    )
  ) |> 
  # creating new column with full site name
  mutate(
    site_full = case_when(
      site_name == "east_channel" ~ "East Channel",
      site_name == "phelps_bridge" ~ "Phelps Bridge",
      site_name == "venoco_bridge" ~ "Venoco Bridge",
      site_name == "other" ~ "Other",
      TRUE ~ site_name
    )
  )
```

## Cleaning water parameter data

```{r}
#| label: cleaning-water-parameter-data

# creating new object from water_parameters
water_parameters_clean <- water_parameters |> 
  # cleaning column names
  clean_names() |> 
  # excluding unnecessary columns
  select(-any_of(c("creator", "editor"))) |> 
  # renaming depth_code as elevation_code
  rename(elevation_code = depth_code) |> 
  # replacing incorrect elevation code from 2021-07-16
  mutate(
    elevation_code = case_when(
      elevation_code == 10 ~ 1,
      TRUE ~ elevation_code
    )
  ) |> 
  # setting elevation code as a factor and ordering levels
  mutate(
    elevation_code = as_factor(elevation_code),
    elevation_code = fct_relevel(
      elevation_code,
      "0", "1", "2", "3", "4", "5", "6"
    )
  )
```

# 3. Join data frames

```{r}
#| label: joining-parameters-and-metadata

# join metadata and water parameters
water_quality <- left_join(
  # left
  x = metadata_clean, 
  # right
  y = water_parameters_clean,
  # shared column
  by = c("global_id" = "parent_global_id")
) |> 
  # selecting relevant columns
  select(
    # information about observation
    global_id, monitoring_datetime, time_of_wq_measurement, 
    monitor_name, weather, site_name, other_site_name, site_full, x, y, 
    
    # water information
    water_surface_elevation, flow, notes, 
    
    # measurement information
    elevation_code, do_mg_l, do_percent_sat, conductivity_u_s_cm,
    salinity_ppt, temperature_c
  ) |> 
  # extracting date from monitoring datetime
  mutate(date = date(monitoring_datetime)) |> 
  # moving date column to appear to the right of monitoring_datetime
  relocate(date, .after = monitoring_datetime)
```

# 4. Filter data for DO figures

```{r}
#| label: filter-do-data

# filter data for dissolved oxygen figures
do_filtered <- water_quality |> 
  # keeping main NCOS sites
  filter(site_full %in% c("East Channel", "Phelps Bridge", "Venoco Bridge")) |> 
  # removing missing dissolved oxygen values
  filter(!is.na(do_mg_l))
```

# 5. Figures

## Figure 1

### Plan

Presenting how dissolved oxygen (DO) differs across sites.

- x-axis: site name
- y-axis: DO in mg/L
- `geom_jitter()`: shows individual DO measurements
- summary point: median DO for each site

```{r}
#| label: figure-1-do-across-sites

ggplot(do_filtered, aes(x = site_full, y = do_mg_l)) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  stat_summary(
    fun = median,
    geom = "point",
    size = 4
  ) +
  labs(
    title = "Dissolved Oxygen Across Sites",
    x = "Site",
    y = "Dissolved oxygen (mg/L)"
  ) +
  theme_bw()
```

## Figure 2

### Plan

Presenting how dissolved oxygen (DO) varies across elevations at each site.

- x-axis: elevation code
- y-axis: DO in mg/L
- `geom_jitter()`: shows individual DO measurements
- summary point: median DO for each elevation code
- panels: site name

```{r}
#| label: figure-2-do-by-elevation

ggplot(do_filtered, aes(x = elevation_code, y = do_mg_l)) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  stat_summary(
    fun = median,
    geom = "point",
    size = 3
  ) +
  facet_wrap(~ site_full) +
  labs(
    title = "Dissolved Oxygen by Elevation Code",
    x = "Elevation code",
    y = "Dissolved oxygen (mg/L)"
  ) +
  theme_bw()
```

## Figure 3

### Plan

Presenting the relationship between dissolved oxygen (DO) and water temperature across sites.

- x-axis: water temperature
- y-axis: DO in mg/L
- color: elevation code
- panels: site name

```{r}
#| label: figure-3-do-temperature

ggplot(
  do_filtered |> filter(!is.na(temperature_c)),
  aes(x = temperature_c, y = do_mg_l, color = elevation_code)
) +
  geom_point(alpha = 0.7) +
  facet_wrap(~ site_full) +
  labs(
    title = "Dissolved Oxygen and Water Temperature",
    x = "Water temperature (°C)",
    y = "Dissolved oxygen (mg/L)",
    color = "Elevation code"
  ) +
  theme_bw()
```