---
title: "Visualization Draft"
author: "Phoebe Dupa, Rebecca Martinez, and Shucan Zhao"
date: today
format:
  pdf: 
    fig-width: 10
    fig-height: 8
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
editor: source
execute: 
  echo: false
  message: false
  warning: false
---

# 1. Set up

## Packages

```{r}
#| label: packages

# general use: data wrangling and plotting
library(tidyverse)

# clean column names
library(janitor)

# file organization
library(here)

# dates and times
library(lubridate)

# non-parametric post-hoc tests and effect sizes
library(rstatix)
```

## Data

```{r}
#| label: data

# water quality survey metadata
metadata <- read_csv(here("data", "NCOS_YSI_Water_Quality_Monitoring_0.csv"))

# water quality parameter measurements
water_parameters <- read_csv(here("data", "YSI_Data_Begin_1.csv"))
```

# 2. Project background

Dissolved oxygen is an important water quality measurement because it affects habitat conditions for aquatic organisms. At NCOS, dissolved oxygen may differ across sites because channels can vary in flow, tidal connection, water residence time, and physical structure.

This project focuses on two questions:

1. How does dissolved oxygen differ across sites?
2. How does dissolved oxygen stratification differ across sites?

These questions are relevant to NCOS because dissolved oxygen helps describe water quality conditions across restored wetland channels. They are also relevant beyond NCOS because low oxygen and stratification can occur in many aquatic systems where depth, flow, and mixing differ across locations.

For this analysis, elevation code is used as a simple way to compare dissolved oxygen at different positions in the water column. Larger differences between bottom and upper measurements suggest stronger dissolved oxygen stratification. Stratification is relevant to the physical processes in this study because it reflects how dissolved oxygen conditions can differ vertically in the water column when water is not fully mixed.

# 3. Clean data

## Cleaning metadata

### Goal

Clean the metadata dataset so it includes survey date, site information, water year, and readable site names.

```{r}
#| label: cleaning-metadata

# create a cleaned metadata dataset
metadata_clean <- metadata |> 
  
  # clean column names so they are easier to use in R
  clean_names() |> 
  
  # remove columns that are not needed for this draft analysis
  select(-any_of(c("object_id", "creator", "editor"))) |> 
  
  # convert monitoring date from character to date-time format
  mutate(monitoring_date = mdy_hms(monitoring_date)) |> 
  
  # rename monitoring date to show that it includes date and time
  rename(monitoring_datetime = monitoring_date) |> 
  
  # create month and year columns from monitoring date-time
  mutate(
    month = month(monitoring_datetime),
    year = year(monitoring_datetime)
  ) |> 
  
  # create water year column
  # water year starts in October, so Oct-Dec are assigned to the next year
  mutate(
    wy = case_when(
      month >= 10 ~ year + 1,
      TRUE ~ year
    )
  ) |> 
  
  # create readable site names for figures and interpretation
  mutate(
    site_full = case_when(
      site_name == "east_channel" ~ "East Channel",
      site_name == "phelps_bridge" ~ "Phelps Bridge",
      site_name == "venoco_bridge" ~ "Venoco Bridge",
      site_name == "other" ~ "Other",
      TRUE ~ site_name
    )
  )
```

## Cleaning water parameter data

### Goal

Clean the water quality parameter dataset so it includes elevation code, elevation labels, and dissolved oxygen measurements.

```{r}
#| label: cleaning-water-parameter-data

# create a cleaned water parameter dataset
water_parameters_clean <- water_parameters |> 
  
  # clean column names so they are easier to use in R
  clean_names() |> 
  
  # remove columns that are not needed for this draft analysis
  select(-any_of(c("creator", "editor"))) |> 
  
  # rename depth_code as elevation_code to match our proposal language
  rename(elevation_code = depth_code) |> 
  
  # correct one known elevation code error from 2021-07-16
  mutate(
    elevation_code = case_when(
      elevation_code == 10 ~ 1,
      TRUE ~ elevation_code
    )
  ) |> 
  
  # create clearer elevation labels for interpretation
  mutate(
    elevation_label = case_when(
      elevation_code == 0 ~ "0: Bottom",
      elevation_code == 1 ~ "1: ~1 ft above bottom / surface",
      elevation_code == 2 ~ "2: ~2 ft above bottom",
      elevation_code == 3 ~ "3: ~3 ft above bottom",
      elevation_code == 4 ~ "4: ~4 ft above bottom",
      elevation_code == 5 ~ "5: ~5 ft above bottom",
      elevation_code == 6 ~ "6: ~6 ft above bottom",
      TRUE ~ as.character(elevation_code)
    )
  ) |> 
  
  # treat elevation code and elevation label as categorical variables
  mutate(
    elevation_code = as_factor(elevation_code),
    elevation_code = fct_relevel(
      elevation_code,
      "0", "1", "2", "3", "4", "5", "6"
    ),
    elevation_label = factor(
      elevation_label,
      levels = c(
        "0: Bottom",
        "1: ~1 ft above bottom / surface",
        "2: ~2 ft above bottom",
        "3: ~3 ft above bottom",
        "4: ~4 ft above bottom",
        "5: ~5 ft above bottom",
        "6: ~6 ft above bottom"
      )
    )
  )
```

# 4. Join data frames

## Water quality = metadata + water parameters

```{r}
#| label: joining-parameters-and-metadata

# join metadata and water parameters
water_quality <- full_join(
  
  # metadata has site and survey information
  x = metadata_clean,
  
  # water_parameters has DO and elevation measurements
  y = water_parameters_clean,
  
  # match GlobalID from metadata with ParentGlobalID from water parameters
  by = c("global_id" = "parent_global_id")
) |> 
  
  # keep columns needed for this project
  select(
    
    # observation information
    global_id,
    monitoring_datetime,
    time_of_wq_measurement,
    monitor_name,
    weather,
    
    # site information
    site_name,
    other_site_name,
    site_full,
    x,
    y,
    
    # date and water year information
    month,
    year,
    wy,
    
    # water condition information
    water_surface_elevation,
    flow,
    notes,
    
    # dissolved oxygen and elevation information
    elevation_code,
    elevation_label,
    do_mg_l,
    do_percent_sat
  ) |> 
  
  # extract date from monitoring datetime
  mutate(date = date(monitoring_datetime)) |> 
  
  # move date column next to monitoring datetime
  relocate(date, .after = monitoring_datetime)
```

# 5. Filter data for dissolved oxygen analysis

### Goal

Create a filtered dissolved oxygen dataset that focuses on the three main NCOS sites, the 2024 and 2025 water years, and the main elevation codes used for the stratification question.

```{r}
#| label: filter-do-data

# create filtered dataset for dissolved oxygen analysis
do_filtered <- water_quality |> 
  
  # remove observations from "Other" sites
  filter(site_full != "Other") |> 
  
  # focus on 2024 and 2025 water years
  filter(wy %in% c(2024, 2025)) |> 
  
  # focus on bottom, ~1 ft above bottom, and ~2 ft above bottom
  filter(elevation_code %in% c("0", "1", "2")) |> 
  
  # keep observations with dissolved oxygen measurements
  filter(!is.na(do_mg_l)) |> 
  
  # remove impossible or extreme DO values that distort the plot scale
  filter(do_mg_l >= 0, do_mg_l <= 30)
```

# 6. Summarize dissolved oxygen

These summaries answer the two main project questions:

1. How does dissolved oxygen differ across sites?
2. How does dissolved oxygen stratification differ across sites?

## Dissolved oxygen by site

```{r}
#| label: summarize-do-by-site

# summarize dissolved oxygen across the three main sites
do_site_summary <- do_filtered |> 
  
  # group observations by site
  group_by(site_full) |> 
  
  # calculate summary statistics for dissolved oxygen
  summarize(
    mean_do = mean(do_mg_l, na.rm = TRUE),
    median_do = median(do_mg_l, na.rm = TRUE),
    q1_do = quantile(do_mg_l, 0.25, na.rm = TRUE),
    q3_do = quantile(do_mg_l, 0.75, na.rm = TRUE),
    sd_do = sd(do_mg_l, na.rm = TRUE),
    var_do = var(do_mg_l, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# make a cleaner table for the PDF
do_site_table <- do_site_summary |> 
  
  # rename columns so the table is easier to read
  rename(
    Site = site_full,
    `Mean DO` = mean_do,
    `Median DO` = median_do,
    `Q1 DO` = q1_do,
    `Q3 DO` = q3_do,
    SD = sd_do,
    Variance = var_do,
    n = n
  )

# show summary table in the rendered PDF
knitr::kable(
  do_site_table,
  digits = 2,
  caption = "Dissolved oxygen summary by site"
)
```

## Dissolved oxygen by site and elevation code

```{r}
#| label: summarize-do-by-site-and-elevation

# summarize dissolved oxygen by site and elevation-code position
do_elevation_summary <- do_filtered |> 
  
  # group observations by site and elevation code
  group_by(site_full, elevation_code) |> 
  
  # calculate summary statistics for dissolved oxygen
  summarize(
    mean_do = mean(do_mg_l, na.rm = TRUE),
    median_do = median(do_mg_l, na.rm = TRUE),
    q1_do = quantile(do_mg_l, 0.25, na.rm = TRUE),
    q3_do = quantile(do_mg_l, 0.75, na.rm = TRUE),
    sd_do = sd(do_mg_l, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# make a cleaner table for the PDF
do_elevation_table <- do_elevation_summary |> 
  
  # rename columns so the table is easier to read
  rename(
    Site = site_full,
    `Elevation code` = elevation_code,
    `Mean DO` = mean_do,
    `Median DO` = median_do,
    `Q1 DO` = q1_do,
    `Q3 DO` = q3_do,
    SD = sd_do,
    n = n
  )

# show summary table in the rendered PDF
knitr::kable(
  do_elevation_table,
  digits = 2,
  caption = "Dissolved oxygen summary by site and elevation code"
)
```

## Stratification difference by site

For this draft, stratification difference is calculated as:

$$\text{Upper median DO} - \text{Bottom median DO}$$

Positive values mean that median DO was higher at the upper measurement position than at the bottom measurement position.

```{r}
#| label: summarize-stratification-difference

# calculate median DO for bottom and upper elevation codes
do_stratification_summary <- do_elevation_summary |> 
  
  # keep bottom and upper measurements only
  filter(elevation_code %in% c("0", "2")) |> 
  
  # create a label for bottom and upper median DO
  mutate(
    position = case_when(
      elevation_code == "0" ~ "bottom_median_do",
      elevation_code == "2" ~ "upper_median_do"
    )
  ) |> 
  
  # keep only columns needed for the stratification calculation
  select(site_full, position, median_do) |> 
  
  # make bottom and upper median DO into separate columns
  pivot_wider(
    names_from = position,
    values_from = median_do
  ) |> 
  
  # calculate upper minus bottom median DO
  mutate(
    stratification_difference = upper_median_do - bottom_median_do
  )

# make a cleaner table for the PDF
do_stratification_table <- do_stratification_summary |> 
  rename(
    Site = site_full,
    `Bottom median DO` = bottom_median_do,
    `Upper median DO` = upper_median_do,
    `Upper - bottom median DO` = stratification_difference
  )

# show stratification difference table
knitr::kable(
  do_stratification_table,
  digits = 2,
  caption = "Dissolved oxygen stratification difference by site"
)
```

# 7. Visualizing

These figures focus on dissolved oxygen. Figure 1 compares dissolved oxygen across sites. Figure 2 compares dissolved oxygen across elevation codes within each site. Figure 3 uses the recommended elevation-profile style to show median dissolved oxygen across elevation codes. Figure 4 adds an exploratory time component by showing median dissolved oxygen through time by elevation code and site.

```{r}
#| label: site-colors

# consistent site colors used across figures
site_colors <- c(
  "East Channel" = "deepskyblue2",
  "Phelps Bridge" = "dodgerblue3",
  "Venoco Bridge" = "darkblue"
)
```

## Figure 1: Dissolved oxygen across sites

**Caption:** Figure 1 shows dissolved oxygen measurements across the three main NCOS sites. Points represent individual observations. Larger black points represent site medians, and vertical ranges show the 1st to 3rd quartile range.

```{r}
#| label: figure-1-do-across-sites
#| fig-width: 10
#| fig-height: 6

# start with filtered dissolved oxygen data
figure_1 <- ggplot(
  data = do_filtered,
  mapping = aes(
    x = site_full,
    y = do_mg_l,
    color = site_full
  )
) +
  # points show individual DO measurements
  geom_point(
    alpha = 0.40,
    size = 1.6,
    position = position_jitter(
      width = 0.2,
      height = 0,
      seed = 193
    )
  ) +
  # black point ranges show median and 1st to 3rd quartiles
  geom_pointrange(
    data = do_site_summary,
    mapping = aes(
      x = site_full,
      y = median_do,
      ymin = q1_do,
      ymax = q3_do
    ),
    inherit.aes = FALSE,
    color = "black",
    size = 0.7
  ) +
  # apply consistent site colors
  scale_color_manual(
    values = site_colors
  ) +
  # keep the y-axis starting at zero
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  # add clear labels
  labs(
    title = "Dissolved Oxygen Across NCOS Sites",
    x = "Site",
    y = "Dissolved oxygen (mg/L)"
  ) +
  # use a simple theme for readability
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

# show figure
figure_1
```

**Interpretation:** Dissolved oxygen varied by site. East Channel had the highest median DO, Phelps Bridge had the lowest median DO, and Venoco Bridge had the widest spread of values.

## Figure 2: Dissolved oxygen stratification by site

**Caption:** Figure 2 shows dissolved oxygen measurements across elevation-code positions within each site. Boxplots show the distribution of DO values, points represent individual observations, and red diamonds represent median DO values.

```{r}
#| label: figure-2-do-stratification-by-site
#| fig-width: 13
#| fig-height: 6

# start with filtered dissolved oxygen data
figure_2 <- ggplot(
  data = do_filtered,
  mapping = aes(
    x = elevation_code,
    y = do_mg_l
  )
) +
  # neutral boxplots show the distribution of DO values at each elevation code
  geom_boxplot(
    outlier.shape = NA,
    width = 0.55,
    alpha = 0.35,
    fill = "gray90",
    color = "gray35"
  ) +
  # colored points show individual DO measurements by site
  geom_point(
    mapping = aes(color = site_full),
    alpha = 0.45,
    size = 1.5,
    position = position_jitter(
      width = 0.12,
      height = 0,
      seed = 193
    )
  ) +
  # red diamonds show median DO for each elevation code
  stat_summary(
    fun = median,
    geom = "point",
    color = "red3",
    size = 4,
    shape = 18
  ) +
  # separate panels by site
  facet_wrap(~ site_full, nrow = 1) +
  # apply consistent site colors to the points
  scale_color_manual(
    values = site_colors
  ) +
  # label elevation codes more clearly
  scale_x_discrete(
    labels = c(
      "0" = "Bottom\n(0)",
      "1" = "Middle\n(1)",
      "2" = "Upper\n(2)"
    )
  ) +
  # keep the y-axis starting at zero
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  # add clear labels
  labs(
    title = "Dissolved Oxygen Stratification by Site",
    x = "Elevation code",
    y = "Dissolved oxygen (mg/L)"
  ) +
  # use a simple theme for readability
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(size = 9),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

# show figure
figure_2
```

**Interpretation:** DO patterns across elevation codes were different by site. East Channel was fairly similar across depths, Phelps Bridge stayed lower overall, and Venoco Bridge showed the clearest increase from bottom to upper measurements.

## Figure 3: Dissolved oxygen elevation profile by site

**Caption:** Figure 3 shows dissolved oxygen elevation profiles for each site. Points represent individual DO observations, red points represent median DO values, and red lines connect median DO values across elevation codes within each site.

```{r}
#| label: figure-3-do-elevation-profile
#| fig-width: 10
#| fig-height: 7

# create profile data with numeric elevation code for plotting
do_profile_df <- do_filtered |> 
  
  # keep only the elevation codes used in this analysis
  filter(elevation_code %in% c("0", "1", "2")) |> 
  
  # convert elevation code to numeric for the profile plot
  mutate(
    elevation_num = as.numeric(as.character(elevation_code))
  )

# create DO elevation profile plot
figure_3 <- ggplot(
  data = do_profile_df,
  mapping = aes(
    x = elevation_num,
    y = do_mg_l,
    group = 1
  )
) +
  # points show individual DO observations
  geom_point(
    shape = 21,
    alpha = 0.35,
    size = 1.5,
    position = position_jitter(
      width = 0.08,
      height = 0,
      seed = 193
    )
  ) +
  # red points show median DO at each elevation code
  stat_summary(
    fun = median,
    geom = "point",
    color = "red3",
    size = 3
  ) +
  # red lines connect median DO values across elevation codes
  stat_summary(
    fun = median,
    geom = "line",
    color = "red3",
    linewidth = 0.8
  ) +
  # separate panels by site
  facet_wrap(
    ~ site_full,
    strip.position = "bottom"
  ) +
  # label elevation codes clearly
  scale_x_continuous(
    breaks = c(0, 1, 2),
    labels = c(
      "Bottom\n(0)",
      "Middle\n(1)",
      "Upper\n(2)"
    )
  ) +
  # keep DO axis starting at zero
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  # flip coordinates so the figure reads more like an elevation profile
  coord_flip() +
  # add clear labels
  labs(
    title = "Dissolved Oxygen Elevation Profile by Site",
    x = "Elevation code",
    y = "Dissolved oxygen (mg/L)"
  ) +
  # use a simple theme for readability
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# show figure
figure_3
```

**Interpretation:** Venoco Bridge had the clearest elevation profile, with median DO increasing from bottom to upper measurements. East Channel and Phelps Bridge showed smaller differences across elevation codes.

## Figure 4: Median dissolved oxygen through time by elevation code

**Caption:** Figure 4 shows median dissolved oxygen through time, separated by elevation code. Points represent median DO for each sampling date and site, and lines connect values within each site.

```{r}
#| label: figure-4-do-through-time
#| fig-width: 10
#| fig-height: 8

# summarize DO by date, site, and elevation code
do_time_summary <- do_filtered |> 
  
  # group observations by sampling date, site, and elevation code
  group_by(date, site_full, elevation_code) |> 
  
  # calculate median DO for each date-site-elevation combination
  summarize(
    median_do = median(do_mg_l, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |> 
  
  # reorder elevation code so upper is at the top and bottom is at the bottom
  mutate(
    panel_elevation = factor(
      elevation_code,
      levels = c("2", "1", "0")
    )
  )

# plot median dissolved oxygen through time
figure_4 <- ggplot(
  data = do_time_summary,
  mapping = aes(
    x = date,
    y = median_do,
    color = site_full,
    group = site_full
  )
) +
  # points show median DO on each sampling date
  geom_point(
    size = 1.7,
    alpha = 0.75
  ) +
  # lines connect median DO values within each site
  geom_line(
    linewidth = 0.6,
    alpha = 0.7
  ) +
  # panel rows show elevation position
  facet_grid(
    rows = vars(panel_elevation),
    labeller = labeller(
      panel_elevation = c(
        "2" = "Upper (2)",
        "1" = "Middle (1)",
        "0" = "Bottom (0)"
      )
    )
  ) +
  # use the same site colors as the other figures
  scale_color_manual(
    values = site_colors,
    name = "Site"
  ) +
  # keep the DO axis starting at zero
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  # add clear labels
  labs(
    title = "Median Dissolved Oxygen Through Time by Elevation Code",
    x = "Sampling date",
    y = "Median dissolved oxygen (mg/L)"
  ) +
  # use a simple theme for readability
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# show figure
figure_4
```

**Interpretation:** Median DO changed through time, but the site pattern was still visible. East Channel was generally higher, Phelps Bridge was generally lower, and Venoco Bridge showed the clearest difference between bottom and upper measurements.

# 8. Statistical analysis

The statistical analyses follow the class workflow for comparing more than two groups: summarize the data, visualize the pattern, run a Kruskal-Wallis test, use Dunn post-hoc tests when the Kruskal-Wallis result is significant, calculate effect size, and interpret the result. The main response variable is dissolved oxygen. The grouping variables are site and elevation code.

## Question 1: How does dissolved oxygen differ across sites?

This analysis tests whether dissolved oxygen differs across East Channel, Phelps Bridge, and Venoco Bridge.

### Kruskal-Wallis test

```{r}
#| label: stats-site-kruskal

# test whether DO differs across the three main sites
do_site_test <- kruskal.test(
  do_mg_l ~ site_full,
  data = do_filtered
)

# show Kruskal-Wallis test result
do_site_test
```

The Kruskal-Wallis test showed that dissolved oxygen differed significantly across sites, $\chi^2$ = 45.67, df = 2, p < 0.001. This means that DO was not distributed the same way across East Channel, Phelps Bridge, and Venoco Bridge.

### Dunn post-hoc test

```{r}
#| label: stats-site-dunn

# compare site pairs after significant Kruskal-Wallis result
do_site_dunn <- do_filtered |> 
  dunn_test(
    do_mg_l ~ site_full,
    p.adjust.method = "holm"
  )

# show Dunn post-hoc test result
do_site_dunn
```

Because the Kruskal-Wallis test was significant, Dunn post-hoc tests were used to compare site pairs. All three site pairs were significantly different after adjustment: East Channel and Phelps Bridge, East Channel and Venoco Bridge, and Phelps Bridge and Venoco Bridge.

### Effect size

```{r}
#| label: stats-site-effect-size

# calculate effect size for the site comparison
do_site_effect_size <- do_filtered |> 
  kruskal_effsize(
    do_mg_l ~ site_full
  )

# show effect size result
do_site_effect_size
```

The effect size for the site comparison was $\eta^2[H]$ = 0.083, which was interpreted as moderate in the test output. This suggests that site explained a moderate amount of variation in DO ranks in this filtered dataset.

### Interpretation

Dissolved oxygen differed across the three main NCOS sites. East Channel had the highest median DO, Phelps Bridge had the lowest median DO, and Venoco Bridge had the widest spread of values. The Dunn post-hoc test showed that all site pairs were significantly different, so DO patterns were not the same across sites.

## Question 2: How does dissolved oxygen stratification differ across sites?

This analysis tests whether dissolved oxygen differs across elevation-code measurement positions within each site. This addresses stratification because elevation code represents the measurement position in the water column.

### Kruskal-Wallis tests within each site

```{r}
#| label: stats-stratification-within-sites-kruskal

# test DO differences across elevation codes within East Channel
east_channel_elevation_test <- kruskal.test(
  do_mg_l ~ elevation_code,
  data = do_filtered |> 
    filter(site_full == "East Channel")
)

# test DO differences across elevation codes within Phelps Bridge
phelps_bridge_elevation_test <- kruskal.test(
  do_mg_l ~ elevation_code,
  data = do_filtered |> 
    filter(site_full == "Phelps Bridge")
)

# test DO differences across elevation codes within Venoco Bridge
venoco_bridge_elevation_test <- kruskal.test(
  do_mg_l ~ elevation_code,
  data = do_filtered |> 
    filter(site_full == "Venoco Bridge")
)

# show within-site Kruskal-Wallis results
east_channel_elevation_test
phelps_bridge_elevation_test
venoco_bridge_elevation_test
```

The within-site Kruskal-Wallis tests showed no significant DO differences across elevation codes at East Channel or Phelps Bridge. Venoco Bridge did show a significant difference across elevation codes, $\chi^2$ = 28.64, df = 2, p < 0.001. This means that Venoco Bridge had the clearest statistical evidence of DO stratification.

### Dunn post-hoc test for Venoco Bridge

```{r}
#| label: stats-venoco-stratification-dunn

# compare elevation code pairs within Venoco Bridge
venoco_bridge_dunn <- do_filtered |> 
  
  # keep only Venoco Bridge because it had a significant Kruskal-Wallis result
  filter(site_full == "Venoco Bridge") |> 
  
  # compare all elevation-code pairs
  dunn_test(
    do_mg_l ~ elevation_code,
    p.adjust.method = "holm"
  )

# show Venoco Bridge Dunn post-hoc result
venoco_bridge_dunn
```

Because Venoco Bridge had a significant Kruskal-Wallis result, Dunn post-hoc tests were used to compare elevation-code pairs at that site. All three pairs were significantly different after adjustment: bottom and middle, bottom and upper, and middle and upper. This supports the pattern of higher median DO from bottom to upper measurements at Venoco Bridge.

### Effect size for Venoco Bridge

```{r}
#| label: stats-venoco-stratification-effect-size

# calculate effect size for elevation-code differences within Venoco Bridge
venoco_bridge_effect_size <- do_filtered |> 
  
  # keep only Venoco Bridge because it showed significant elevation-code differences
  filter(site_full == "Venoco Bridge") |> 
  
  # calculate Kruskal-Wallis effect size
  kruskal_effsize(
    do_mg_l ~ elevation_code
  )

# show Venoco Bridge effect size result
venoco_bridge_effect_size
```

The effect size for the Venoco Bridge elevation-code comparison was $\eta^2[H]$ = 0.105, which was interpreted as moderate in the test output. This suggests that elevation code explained a moderate amount of variation in DO ranks at Venoco Bridge.

### Interpretation

Statistical evidence for DO stratification was clearest at Venoco Bridge. East Channel and Phelps Bridge did not show significant DO differences across elevation codes, while Venoco Bridge did. At Venoco Bridge, median DO was lower at the bottom and higher at the middle and upper positions.

# 9. Advanced elective plan

For the advanced elective, the group plans to create a physical version of a dissolved oxygen stratification model. We will use three clear containers to represent East Channel, Phelps Bridge, and Venoco Bridge. Each container will show bottom, middle, and upper elevation-code positions.

Blue beads will represent water in each container. The beads have already been separated into different shades of blue so each site can be visually distinct. Cotton from cotton balls will be layered more densely or sparsely to represent relative dissolved oxygen concentration at each elevation position. The visual mock-up below was created to help guide the physical model.

## Elective schedule

By Week 9, we separated the blue beads by color, created the visual mock-up, and finalized the general plan for representing the three sites.

During Week 10, we will finish the final layering plan for each tube, add markings to show bottom, middle, and upper elevation-code positions, and assemble the physical model.

By the final presentation, the three containers will be completed and used to explain how dissolved oxygen stratification differs across the NCOS sites.

```{r}
#| label: elective-visual-mock-up
#| fig-width: 12
#| fig-height: 6

# reorder elevation code so bottom is at the bottom and upper is at the top
do_plot_elective <- do_filtered |> 
  mutate(
    panel_elevation = factor(
      elevation_code,
      levels = c("0", "1", "2")
    )
  )

# set site order for panels
site_levels <- c("East Channel", "Phelps Bridge", "Venoco Bridge")

# create one row per panel for the colored backgrounds
panel_background <- do_plot_elective |> 
  distinct(site_full) |> 
  rename(panel_site = site_full) |> 
  mutate(panel_site = factor(panel_site, levels = site_levels))

# create focal points so each site appears only in its own panel
highlight_focal <- do_plot_elective |> 
  mutate(
    panel_site = factor(site_full, levels = site_levels)
  )

# make visual mock-up for the physical elective idea
elective_mock_up <- ggplot() +
  
  # blue background represents water in each site container
  geom_rect(
    data = panel_background,
    aes(
      xmin = -Inf,
      xmax = Inf,
      ymin = -Inf,
      ymax = Inf,
      fill = panel_site
    ),
    inherit.aes = FALSE,
    alpha = 0.45,
    show.legend = FALSE
  ) +
  
  # white points show dissolved oxygen observations for each site
  geom_point(
    data = highlight_focal,
    aes(
      x = do_mg_l,
      y = panel_elevation
    ),
    color = "white",
    alpha = 0.85,
    size = 1.9,
    position = position_jitter(
      width = 0,
      height = 0.12,
      seed = 193
    )
  ) +
  
  # one panel for each site
  facet_wrap(~ panel_site, nrow = 1) +
  
  # use the same site colors as the other figures
  scale_fill_manual(
    values = site_colors
  ) +
  
  # keep the DO axis starting at zero
  scale_x_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  
  # relabel elevation code on the y-axis
  scale_y_discrete(
    labels = c(
      "0" = "Bottom (0)",
      "1" = "Middle (1)",
      "2" = "Upper (2)"
    )
  ) +
  
  # add clear labels
  labs(
    title = "Physical Model Mock-Up for Dissolved Oxygen Stratification",
    x = "Dissolved oxygen (mg/L)",
    y = "Elevation code"
  ) +
  
  # use a simple theme for readability
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "white"),
    panel.border = element_rect(color = "black", fill = NA),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

# show elective mock-up
elective_mock_up
```

The physical model will use the same basic idea as the mock-up: each site will have a visible water-column structure, and the bottom, middle, and upper positions will be represented separately. Based on the data, Venoco Bridge should show the clearest difference between bottom and upper measurements, while East Channel and Phelps Bridge should show smaller vertical differences.



# 10. Overall interpretation

Overall, dissolved oxygen differed across the three main NCOS sites, and statistical evidence for stratification was clearest at Venoco Bridge. East Channel had the highest median DO, Phelps Bridge had the lowest median DO, and Venoco Bridge showed the clearest increase from bottom to upper measurements. These results suggest that DO conditions and stratification patterns were not the same across the three sites.