---
title: "Visualization Draft"
author: "Phoebe Dupa, Rebecca Martinez, and Shucan Zhao"
date: "Spring 2026"
format:
  pdf: 
    fig-width: 10
    fig-height: 8
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
editor: source
execute: 
  message: false
  warning: false
---

# 1. Set up

## Packages

```{r}
#| label: packages
#| message: false

# general use: data wrangling and plotting
library(tidyverse)

# clean column names
library(janitor)

# file organization
library(here)

# working with dates and times
library(lubridate)
```

## Plot settings

```{r}
#| label: plot-settings

# set a common theme for all plots
theme_set(
  theme_bw() +
    theme(
      text = element_text(size = 12),
      plot.title.position = "plot",
      legend.position = "right"
    )
)
```

## Data

```{r data}

# weather data from NOAA weather station
weather <- read_csv(here("data", "NOAA-weather-data.csv"))

# water quality survey metadata
metadata <- read_csv(here("data", "NCOS_YSI_Water_Quality_Monitoring_0.csv"))

# water quality parameter measurements
water_parameters <- read_csv(here("data", "YSI_Data_Begin_1.csv"))
```

# 2. Clean data

## Cleaning weather data

```{r cleaning-weather-data}


# create a cleaned weather dataset
weather_clean <- weather |> 
  
  # clean column names
  clean_names() |> 
  
  # make sure date is understood as a date
  # DATE is stored as a character vector when read in, so mdy() converts it
  mutate(date = mdy(date)) |> 
  
  # keep columns that may be useful for interpretation
  select(date, prcp, tmax, tmin) |> 
  
  # create month and year columns from date
  mutate(
    month = month(date),
    year = year(date)
  ) |> 
  
  # create water year column
  # water year starts in October
  mutate(
    wy = case_when(
      month >= 10 ~ year + 1,
      TRUE ~ year
    )
  )
```

## Cleaning metadata

```{r}
#| label: cleaning-metadata

# create a cleaned metadata dataset
metadata_clean <- metadata |> 
  
  # clean column names
  clean_names() |> 
  
  # remove columns that are not needed for this project
  select(-any_of(c("object_id", "creator", "editor"))) |> 
  
  # make sure monitoring_date is understood as date-time
  # Monitoring Date is stored as a character vector when read in
  mutate(monitoring_date = mdy_hms(monitoring_date)) |> 
  
  # rename monitoring date column to show that it includes date and time
  rename(monitoring_datetime = monitoring_date) |> 
  
  # create month and year columns from monitoring_datetime
  mutate(
    month = month(monitoring_datetime),
    year = year(monitoring_datetime)
  ) |> 
  
  # create water year column
  # water year starts in October
  mutate(
    wy = case_when(
      month >= 10 ~ year + 1,
      TRUE ~ year
    )
  ) |> 
  
  # create readable site names for plotting and interpretation
  mutate(
    site_full = case_when(
      site_name == "east_channel" ~ "East Channel",
      site_name == "phelps_bridge" ~ "Phelps Bridge",
      site_name == "venoco_bridge" ~ "Venoco Bridge",
      site_name == "other" ~ "Other",
      TRUE ~ site_name
    )
  )
```

## Cleaning water parameter data

```{r}
#| label: cleaning-water-parameter-data

# create a cleaned water parameter dataset
water_parameters_clean <- water_parameters |> 
  
  # clean column names
  clean_names() |> 
  
  # remove columns that are not needed for this project
  select(-any_of(c("creator", "editor"))) |> 
  
  # rename depth_code as elevation_code to match our proposal language
  rename(elevation_code = depth_code) |> 
  
  # replace incorrect elevation code from 2021-07-16
  mutate(
    elevation_code = case_when(
      elevation_code == 10 ~ 1,
      TRUE ~ elevation_code
    )
  ) |> 
  
  # treat elevation code as categorical
  mutate(
    elevation_code = as_factor(elevation_code),
    
    # order elevation codes from 0 to 6
    elevation_code = fct_relevel(
      elevation_code,
      "0", "1", "2", "3", "4", "5", "6"
    )
  )
```

# 3. Join data frames

## Water quality = metadata + water parameters

```{r}
#| label: joining-parameters-and-metadata

# join metadata and water parameters
water_quality <- full_join(
  
  # metadata has site and survey information
  x = metadata_clean,
  
  # water_parameters has DO, salinity, temperature, and elevation measurements
  y = water_parameters_clean,
  
  # match GlobalID from metadata with ParentGlobalID from water parameters
  by = c("global_id" = "parent_global_id")
) |> 
  
  # keep columns needed for this project
  select(
    
    # information about observation
    global_id,
    monitoring_datetime,
    time_of_wq_measurement,
    monitor_name,
    weather,
    site_name,
    other_site_name,
    site_full,
    x,
    y,
    month,
    year,
    wy,
    
    # water information
    water_surface_elevation,
    flow,
    notes,
    
    # water quality measurements
    elevation_code,
    do_mg_l,
    do_percent_sat,
    conductivity_u_s_cm,
    salinity_ppt,
    temperature_c
  ) |> 
  
  # extract date from monitoring datetime
  mutate(date = date(monitoring_datetime)) |> 
  
  # move date column next to monitoring datetime
  relocate(date, .after = monitoring_datetime)
```

## Water quality weather = water quality + weather

```{r}
#| label: joining-water-quality-and-weather

# join water quality data with weather data by date
water_quality_weather <- left_join(
  
  # water_quality has site, DO, temperature, salinity, and elevation data
  x = water_quality,
  
  # weather_clean has precipitation and air temperature data
  y = weather_clean,
  
  # shared column
  by = "date"
)
```

# 4. Filter data for dissolved oxygen analysis

```{r}
#| label: filter-do-data

# create filtered dataset for dissolved oxygen analysis
do_filtered <- water_quality_weather |> 
  
  # keep only the three main NCOS sites from our proposal
  filter(site_full != "Other") |> 
  
  # keep observations with dissolved oxygen measurements
  filter(!is.na(do_mg_l))
```

# 5. Summarize dissolved oxygen

## Dissolved oxygen by site

```{r}
#| label: summarize-do-by-site

# summarize dissolved oxygen across sites
do_site_summary <- do_filtered |> 
  
  # group observations by site
  group_by(site_full) |> 
  
  # calculate summary statistics for dissolved oxygen
  summarize(
    mean_do = mean(do_mg_l, na.rm = TRUE),
    median_do = median(do_mg_l, na.rm = TRUE),
    sd_do = sd(do_mg_l, na.rm = TRUE),
    var_do = var(do_mg_l, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# show summary table
do_site_summary
```

## Dissolved oxygen by site and elevation code

```{r}
#| label: summarize-do-by-elevation

# summarize dissolved oxygen by site and elevation code
do_elevation_summary <- do_filtered |> 
  
  # remove observations without elevation code
  filter(!is.na(elevation_code)) |> 
  
  # group observations by site and elevation code
  group_by(site_full, elevation_code) |> 
  
  # calculate summary statistics for dissolved oxygen
  summarize(
    mean_do = mean(do_mg_l, na.rm = TRUE),
    median_do = median(do_mg_l, na.rm = TRUE),
    sd_do = sd(do_mg_l, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# show summary table
do_elevation_summary
```

# 6. Figures

## Figure 1: Dissolved oxygen across sites

### Plan

This figure compares dissolved oxygen measurements across the three main NCOS sites.

- x-axis: `site_full`
- y-axis: `do_mg_l`
- `geom_jitter()`: shows individual DO measurements
- `stat_summary()`: shows median DO for each site

```{r}
#| label: figure-1-do-across-sites

# compare dissolved oxygen values across sites
ggplot(
  data = do_filtered,
  mapping = aes(
    x = site_full,
    y = do_mg_l
  )
) +
  
  # jitter points represent individual dissolved oxygen observations
  geom_jitter(
    width = 0.2,
    height = 0,
    alpha = 0.5
  ) +
  
  # larger point shows median dissolved oxygen for each site
  stat_summary(
    fun = median,
    geom = "point",
    size = 4
  ) +
  
  # relabel axes and add title
  labs(
    title = "Dissolved Oxygen Across Sites",
    x = "Site",
    y = "Dissolved oxygen (mg/L)"
  )
```

## Figure 2: Dissolved oxygen and water temperature

### Plan

This figure compares dissolved oxygen with water temperature across sites.

- x-axis: `temperature_c`
- y-axis: `do_mg_l`
- color: `elevation_code`
- panels: `site_full`
- `geom_point()`: shows individual observations

```{r}
#| label: figure-2-do-temperature

# compare dissolved oxygen and water temperature across sites
ggplot(
  data = do_filtered |> 
    filter(!is.na(temperature_c), !is.na(elevation_code)),
  mapping = aes(
    x = temperature_c,
    y = do_mg_l,
    color = elevation_code
  )
) +
  
  # points represent individual observations
  geom_point(alpha = 0.7) +
  
  # create separate panel for each site
  facet_wrap(~ site_full) +
  
  # relabel axes, legend, and title
  labs(
    title = "Dissolved Oxygen and Water Temperature",
    x = "Water temperature (°C)",
    y = "Dissolved oxygen (mg/L)",
    color = "Elevation code"
  )
```

## Figure 3: Dissolved oxygen elevation profile

### Plan

This figure compares dissolved oxygen measurements across elevation codes at each site.

- x-axis: `elevation_code`
- y-axis: `do_mg_l`
- panels: `site_full`
- `geom_jitter()`: shows individual observations
- `stat_summary()`: shows median DO at each elevation code
- `stat_summary(geom = "line")`: connects median DO values
- `coord_flip()`: makes the figure read more like an elevation profile

```{r}
#| label: figure-3-do-elevation-profile

# compare dissolved oxygen across elevation codes within each site
ggplot(
  data = do_filtered |> 
    filter(!is.na(elevation_code)),
  mapping = aes(
    x = elevation_code,
    y = do_mg_l,
    group = 1
  )
) +
  
  # jitter points represent individual DO observations
  geom_jitter(
    width = 0.15,
    height = 0,
    alpha = 0.5
  ) +
  
  # larger points show median DO for each elevation code
  stat_summary(
    fun = median,
    geom = "point",
    size = 3
  ) +
  
  # line connects median DO values across elevation codes
  stat_summary(
    fun = median,
    geom = "line"
  ) +
  
  # create separate panel for each site
  facet_wrap(~ site_full) +
  
  # flip coordinates to make figure read more like an elevation profile
  coord_flip() +
  
  # relabel axes and add title
  labs(
    title = "Dissolved Oxygen at Different Elevations",
    x = "Elevation code",
    y = "Dissolved oxygen (mg/L)"
  )
```

# 7. Interpretation notes

These summaries and figures will help us answer our main project questions.

The site summary and Figure 1 will help compare dissolved oxygen across East Channel, Phelps Bridge, and Venoco Bridge. The elevation summary and Figure 3 will help us look for vertical differences in dissolved oxygen within each site. If dissolved oxygen changes clearly across elevation codes, that may suggest possible stratification. Figure 2 will help us explore whether water temperature may help explain dissolved oxygen patterns.

This is still a draft analysis, so filtering choices and figure details may be adjusted as we learn more about the dataset.