---
title: "Visualization Draft"
author: "Phoebe Dupa, Rebecca Martinez, and Shucan Zhao"
date: "Spring 2026"
format:
  pdf: 
    fig-width: 10
    fig-height: 8
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
editor: source
execute: 
  echo: false
  message: false
  warning: false
---

# 1. Set up

## Packages

```{r}
#| label: packages

# general use: data wrangling and plotting
library(tidyverse)

# clean column names
library(janitor)

# file organization
library(here)

# working with dates and times
library(lubridate)
```

## Plot settings

```{r}
#| label: plot-settings

# set a common theme for all plots
theme_set(
  theme_bw() +
    theme(
      text = element_text(size = 12),
      plot.title.position = "plot",
      legend.position = "right"
    )
)
```

## Data

```{r}
#| label: data

# weather data from NOAA weather station
weather <- read_csv(here("data", "NOAA-weather-data.csv"))

# water quality survey metadata
metadata <- read_csv(here("data", "NCOS_YSI_Water_Quality_Monitoring_0.csv"))

# water quality parameter measurements
water_parameters <- read_csv(here("data", "YSI_Data_Begin_1.csv"))
```

# 2. Clean data

## Cleaning weather data

### Goal

Clean weather dataset with date, precipitation, maximum temperature, minimum temperature, month, year, and water year.

```{r}
#| label: cleaning-weather-data

# create a cleaned weather dataset
weather_clean <- weather |> 
  
  # clean column names so they are easier to use in R
  clean_names() |> 
  
  # convert date from character to date format
  mutate(date = mdy(date)) |> 
  
  # keep columns that may be useful for interpretation
  select(date, prcp, tmax, tmin) |> 
  
  # create month and year columns from date
  mutate(
    month = month(date),
    year = year(date)
  ) |> 
  
  # create water year column
  # water year starts in October, so Oct-Dec are assigned to the next year
  mutate(
    wy = case_when(
      month >= 10 ~ year + 1,
      TRUE ~ year
    )
  )
```

## Cleaning metadata

### Goal

Clean metadata dataset with survey date, site information, water year, and readable site names.

```{r}
#| label: cleaning-metadata

# create a cleaned metadata dataset
metadata_clean <- metadata |> 
  
  # clean column names so they are easier to use in R
  clean_names() |> 
  
  # remove columns that are not needed for this draft analysis
  select(-any_of(c("object_id", "creator", "editor"))) |> 
  
  # convert monitoring date from character to date-time format
  mutate(monitoring_date = mdy_hms(monitoring_date)) |> 
  
  # rename monitoring date to show that it includes date and time
  rename(monitoring_datetime = monitoring_date) |> 
  
  # create month and year columns from monitoring date-time
  mutate(
    month = month(monitoring_datetime),
    year = year(monitoring_datetime)
  ) |> 
  
  # create water year column
  # water year starts in October, so Oct-Dec are assigned to the next year
  mutate(
    wy = case_when(
      month >= 10 ~ year + 1,
      TRUE ~ year
    )
  ) |> 
  
  # create readable site names for figures and interpretation
  mutate(
    site_full = case_when(
      site_name == "east_channel" ~ "East Channel",
      site_name == "phelps_bridge" ~ "Phelps Bridge",
      site_name == "venoco_bridge" ~ "Venoco Bridge",
      site_name == "other" ~ "Other",
      TRUE ~ site_name
    )
  )
```

## Cleaning water parameter data

### Goal

Clean water quality parameter dataset with elevation code, dissolved oxygen, temperature, salinity, and conductivity.

```{r}
#| label: cleaning-water-parameter-data

# create a cleaned water parameter dataset
water_parameters_clean <- water_parameters |> 
  
  # clean column names so they are easier to use in R
  clean_names() |> 
  
  # remove columns that are not needed for this draft analysis
  select(-any_of(c("creator", "editor"))) |> 
  
  # rename depth_code as elevation_code to match our proposal language
  rename(elevation_code = depth_code) |> 
  
  # correct one known elevation code error from 2021-07-16
  mutate(
    elevation_code = case_when(
      elevation_code == 10 ~ 1,
      TRUE ~ elevation_code
    )
  ) |> 
  
  # treat elevation code as categorical and order levels
  mutate(
    elevation_code = as_factor(elevation_code),
    elevation_code = fct_relevel(
      elevation_code,
      "0", "1", "2", "3", "4", "5", "6"
    )
  )
```

# 3. Join data frames

## Water quality = metadata + water parameters

```{r}
#| label: joining-parameters-and-metadata

# join metadata and water parameters
water_quality <- full_join(
  
  # metadata has site and survey information
  x = metadata_clean,
  
  # water_parameters has DO, salinity, temperature, and elevation measurements
  y = water_parameters_clean,
  
  # match GlobalID from metadata with ParentGlobalID from water parameters
  by = c("global_id" = "parent_global_id")
) |> 
  
  # keep columns needed for this project
  select(
    
    # information about observation
    global_id,
    monitoring_datetime,
    time_of_wq_measurement,
    monitor_name,
    weather,
    site_name,
    other_site_name,
    site_full,
    x,
    y,
    month,
    year,
    wy,
    
    # water information
    water_surface_elevation,
    flow,
    notes,
    
    # water quality measurements
    elevation_code,
    do_mg_l,
    do_percent_sat,
    conductivity_u_s_cm,
    salinity_ppt,
    temperature_c
  ) |> 
  
  # extract date from monitoring datetime
  mutate(date = date(monitoring_datetime)) |> 
  
  # move date column next to monitoring datetime
  relocate(date, .after = monitoring_datetime)
```

## Water quality weather = water quality + weather

```{r}
#| label: joining-water-quality-and-weather

# join water quality data with weather data by date
water_quality_weather <- left_join(
  
  # water_quality has site, DO, temperature, salinity, and elevation data
  x = water_quality,
  
  # weather_clean has precipitation and air temperature data
  y = weather_clean,
  
  # shared column
  by = "date"
)
```

# 4. Filter data for dissolved oxygen analysis

### Goal

Create a filtered dissolved oxygen dataset that focuses on the three main NCOS sites.

```{r}
#| label: filter-do-data

# create filtered dataset for dissolved oxygen analysis
do_filtered <- water_quality_weather |> 
  
  # remove observations from "Other" sites
  filter(site_full != "Other") |> 
  
  # keep observations with dissolved oxygen measurements
  filter(!is.na(do_mg_l))
```

# 5. Summarize dissolved oxygen

## Dissolved oxygen by site

```{r}
#| label: summarize-do-by-site

# summarize dissolved oxygen across sites
do_site_summary <- do_filtered |> 
  
  # group observations by site
  group_by(site_full) |> 
  
  # calculate summary statistics for dissolved oxygen
  summarize(
    mean_do = mean(do_mg_l, na.rm = TRUE),
    median_do = median(do_mg_l, na.rm = TRUE),
    sd_do = sd(do_mg_l, na.rm = TRUE),
    var_do = var(do_mg_l, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

# show summary table
do_site_summary
```

