---
title: "Visualization Draft"
author: "Phoebe Dupa, Rebecca Martinez, and Shucan Zhao"
date: today
format:
  pdf: 
    fig-width: 10
    fig-height: 8
    include-in-header:
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere,commandchars=\\\{\}}
editor: source
execute: 
  echo: false
  message: false
  warning: false
---

# 1. Set up

## Packages

```{r}
#| label: packages

# general use: data wrangling and plotting
library(tidyverse)

# clean column names
library(janitor)

# file organization
library(here)

# dates and times
library(lubridate)

# non-parametric post-hoc tests and effect sizes
library(rstatix)

```

## Data

```{r}
#| label: data

# water quality survey metadata
metadata <- read_csv(here("data", "NCOS_YSI_Water_Quality_Monitoring_0.csv"))

# water quality parameter measurements
water_parameters <- read_csv(here("data", "YSI_Data_Begin_1.csv"))
```

# 2. Clean data

## Cleaning metadata

### Goal

Clean the metadata dataset so it includes survey date, site information, water year, and readable site names.

```{r}
#| label: cleaning-metadata

# create a cleaned metadata dataset
metadata_clean <- metadata |> 
  
  # clean column names so they are easier to use in R
  clean_names() |> 
  
  # remove columns that are not needed for this draft analysis
  select(-any_of(c("object_id", "creator", "editor"))) |> 
  
  # convert monitoring date from character to date-time format
  mutate(monitoring_date = mdy_hms(monitoring_date)) |> 
  
  # rename monitoring date to show that it includes date and time
  rename(monitoring_datetime = monitoring_date) |> 
  
  # create month and year columns from monitoring date-time
  mutate(
    month = month(monitoring_datetime),
    year = year(monitoring_datetime)
  ) |> 
  
  # create water year column
  # water year starts in October, so Oct-Dec are assigned to the next year
  mutate(
    wy = case_when(
      month >= 10 ~ year + 1,
      TRUE ~ year
    )
  ) |> 
  
  # create readable site names for figures and interpretation
  mutate(
    site_full = case_when(
      site_name == "east_channel" ~ "East Channel",
      site_name == "phelps_bridge" ~ "Phelps Bridge",
      site_name == "venoco_bridge" ~ "Venoco Bridge",
      site_name == "other" ~ "Other",
      TRUE ~ site_name
    )
  )
```

## Cleaning water parameter data

### Goal

Clean the water quality parameter dataset so it includes elevation code, elevation labels, and dissolved oxygen measurements.

```{r}
#| label: cleaning-water-parameter-data

# create a cleaned water parameter dataset
water_parameters_clean <- water_parameters |> 
  
  # clean column names so they are easier to use in R
  clean_names() |> 
  
  # remove columns that are not needed for this draft analysis
  select(-any_of(c("creator", "editor"))) |> 
  
  # rename depth_code as elevation_code to match our proposal language
  rename(elevation_code = depth_code) |> 
  
  # correct one known elevation code error from 2021-07-16
  mutate(
    elevation_code = case_when(
      elevation_code == 10 ~ 1,
      TRUE ~ elevation_code
    )
  ) |> 
  
  # create clearer elevation labels for interpretation
  mutate(
    elevation_label = case_when(
      elevation_code == 0 ~ "0: Bottom",
      elevation_code == 1 ~ "1: ~1 ft above bottom / surface",
      elevation_code == 2 ~ "2: ~2 ft above bottom",
      elevation_code == 3 ~ "3: ~3 ft above bottom",
      elevation_code == 4 ~ "4: ~4 ft above bottom",
      elevation_code == 5 ~ "5: ~5 ft above bottom",
      elevation_code == 6 ~ "6: ~6 ft above bottom",
      TRUE ~ as.character(elevation_code)
    )
  ) |> 
  
  # treat elevation code and elevation label as categorical variables
  mutate(
    elevation_code = as_factor(elevation_code),
    elevation_code = fct_relevel(
      elevation_code,
      "0", "1", "2", "3", "4", "5", "6"
    ),
    elevation_label = factor(
      elevation_label,
      levels = c(
        "0: Bottom",
        "1: ~1 ft above bottom / surface",
        "2: ~2 ft above bottom",
        "3: ~3 ft above bottom",
        "4: ~4 ft above bottom",
        "5: ~5 ft above bottom",
        "6: ~6 ft above bottom"
      )
    )
  )
```

# 3. Join data frames

## Water quality = metadata + water parameters

```{r}
#| label: joining-parameters-and-metadata

# join metadata and water parameters
water_quality <- full_join(
  
  # metadata has site and survey information
  x = metadata_clean,
  
  # water_parameters has DO and elevation measurements
  y = water_parameters_clean,
  
  # match GlobalID from metadata with ParentGlobalID from water parameters
  by = c("global_id" = "parent_global_id")
) |> 
  
  # keep columns needed for this project
  select(
    
    # observation information
    global_id,
    monitoring_datetime,
    time_of_wq_measurement,
    monitor_name,
    weather,
    
    # site information
    site_name,
    other_site_name,
    site_full,
    x,
    y,
    
    # date and water year information
    month,
    year,
    wy,
    
    # water condition information
    water_surface_elevation,
    flow,
    notes,
    
    # dissolved oxygen and elevation information
    elevation_code,
    elevation_label,
    do_mg_l,
    do_percent_sat
  ) |> 
  
  # extract date from monitoring datetime
  mutate(date = date(monitoring_datetime)) |> 
  
  # move date column next to monitoring datetime
  relocate(date, .after = monitoring_datetime)
```

# 4. Filter data for dissolved oxygen analysis

### Goal

Create a filtered dissolved oxygen dataset that focuses on the three main NCOS sites, the 2024 and 2025 water years, and the main elevation codes used for the stratification question.

```{r}
#| label: filter-do-data

# create filtered dataset for dissolved oxygen analysis
do_filtered <- water_quality |> 
  
  # remove observations from "Other" sites
  filter(site_full != "Other") |> 
  
  # focus on 2024 and 2025 water years
  filter(wy %in% c(2024, 2025)) |> 
  
  # focus on bottom, ~1 ft above bottom, and ~2 ft above bottom
  filter(elevation_code %in% c("0", "1", "2")) |> 
  
  # keep observations with dissolved oxygen measurements
  filter(!is.na(do_mg_l)) |> 
  
  # remove impossible or extreme DO values that distort the plot scale
  filter(do_mg_l >= 0, do_mg_l <= 30)
```

# 5. Summarize dissolved oxygen

These summaries answer the two main project questions:

1. How does dissolved oxygen differ across sites?
2. How does dissolved oxygen stratification differ across sites?

## Dissolved oxygen by site

```{r}
#| label: summarize-do-by-site

# summarize dissolved oxygen across the three main sites
do_site_summary <- do_filtered |> 
  
  # group observations by site
  group_by(site_full) |> 
  
  # calculate summary statistics for dissolved oxygen
  summarize(
    mean_do = mean(do_mg_l, na.rm = TRUE),
    median_do = median(do_mg_l, na.rm = TRUE),
    sd_do = sd(do_mg_l, na.rm = TRUE),
    var_do = var(do_mg_l, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |> 
  
  # rename columns so the table is easier to read
  rename(
    Site = site_full,
    `Mean DO` = mean_do,
    `Median DO` = median_do,
    SD = sd_do,
    Variance = var_do,
    n = n
  )

# show summary table in the rendered PDF
knitr::kable(
  do_site_summary,
  digits = 2,
  caption = "Dissolved oxygen summary by site"
)
```

## Dissolved oxygen by site and elevation code

```{r}
#| label: summarize-do-by-site-and-elevation

# summarize dissolved oxygen by site and elevation-code position
do_elevation_summary <- do_filtered |> 
  
  # group observations by site and elevation code
  group_by(site_full, elevation_code) |> 
  
  # calculate summary statistics for dissolved oxygen
  summarize(
    mean_do = mean(do_mg_l, na.rm = TRUE),
    median_do = median(do_mg_l, na.rm = TRUE),
    sd_do = sd(do_mg_l, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |> 
  
  # rename columns so the table is easier to read
  rename(
    Site = site_full,
    `Elevation code` = elevation_code,
    `Mean DO` = mean_do,
    `Median DO` = median_do,
    SD = sd_do,
    n = n
  )

# show summary table in the rendered PDF
knitr::kable(
  do_elevation_summary,
  digits = 2,
  caption = "Dissolved oxygen summary by site and elevation code"
)
```

# 6. Visualizing

These figures focus on dissolved oxygen. Figure 1 follows the original proposed site comparison. Figure 2 follows the original proposed elevation-code stratification figure. Figure 3 replaces the scrapped temperature figure with a highlighted DO stratification figure. Figure 4 adds an exploratory time component by showing dissolved oxygen through time by site and elevation code.

```{r}
#| label: site-colors

# consistent site colors used across figures
site_colors <- c(
  "East Channel" = "deepskyblue2",
  "Phelps Bridge" = "dodgerblue3",
  "Venoco Bridge" = "darkblue"
)
```

## Figure 1: Dissolved oxygen across sites

**Caption:** Figure 1 shows dissolved oxygen measurements across the three main NCOS sites. Points represent individual observations, and red diamonds represent site medians.

```{r}
#| label: figure-1-do-across-sites
#| fig-width: 10
#| fig-height: 6

# start with filtered dissolved oxygen data
figure_1 <- ggplot(
  data = do_filtered,
  mapping = aes(
    x = site_full,
    y = do_mg_l,
    color = site_full
  )
) +
  # points show individual DO measurements
  geom_point(
    alpha = 0.45,
    size = 1.8,
    position = position_jitter(
      width = 0.2,
      height = 0,
      seed = 193
    )
  ) +
  # red diamonds show median DO for each site
  stat_summary(
    fun = median,
    geom = "point",
    color = "red3",
    size = 5,
    shape = 18
  ) +
  # apply consistent site colors
  scale_color_manual(
    values = site_colors
  ) +
  # keep the y-axis starting at zero
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  # add clear labels
  labs(
    title = "Dissolved Oxygen Across NCOS Sites",
    x = "Site",
    y = "Dissolved oxygen (mg/L)"
  ) +
  # use a simple theme for readability
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

# show figure
figure_1
```

**Interpretation:** Dissolved oxygen varied by site. East Channel had the highest median DO, Phelps Bridge had the lowest, and Venoco Bridge had the widest range of values.

## Figure 2: Dissolved oxygen stratification by site

**Caption:** Figure 2 shows dissolved oxygen measurements across elevation-code positions within each site. Boxplots show the distribution of DO values, points represent individual observations, and red diamonds represent median DO values.

```{r}
#| label: figure-2-do-stratification-by-site
#| fig-width: 13
#| fig-height: 6

# start with filtered dissolved oxygen data
figure_2 <- ggplot(
  data = do_filtered,
  mapping = aes(
    x = elevation_code,
    y = do_mg_l
  )
) +
  # neutral boxplots show the distribution of DO values at each elevation code
  geom_boxplot(
    outlier.shape = NA,
    width = 0.55,
    alpha = 0.35,
    fill = "gray90",
    color = "gray35"
  ) +
  # colored points show individual DO measurements by site
  geom_point(
    mapping = aes(color = site_full),
    alpha = 0.45,
    size = 1.5,
    position = position_jitter(
      width = 0.12,
      height = 0,
      seed = 193
    )
  ) +
  # red diamonds show median DO for each elevation code
  stat_summary(
    fun = median,
    geom = "point",
    color = "red3",
    size = 4,
    shape = 18
  ) +
  # separate panels by site
  facet_wrap(~ site_full, nrow = 1) +
  # apply consistent site colors to the points
  scale_color_manual(
    values = site_colors
  ) +
  # label elevation codes more clearly
  scale_x_discrete(
    labels = c(
      "0" = "Bottom\n(0)",
      "1" = "Middle\n(1)",
      "2" = "Upper\n(2)"
    )
  ) +
  # keep the y-axis starting at zero
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  # add clear labels
  labs(
    title = "Dissolved Oxygen Stratification by Site",
    x = "Elevation code",
    y = "Dissolved oxygen (mg/L)"
  ) +
  # use a simple theme for readability
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(size = 9),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

# show figure
figure_2
```

**Interpretation:** DO patterns across elevation codes were different by site. East Channel was fairly similar across depths, Phelps Bridge stayed lower overall, and Venoco Bridge showed the clearest increase from bottom to upper measurements.

## Figure 3: Dissolved oxygen stratification by focal site

**Caption:** Figure 3 highlights dissolved oxygen measurements across elevation-code positions for each focal site. In each panel, white points show the focal site and gray points show the other sites for comparison. Elevation code is shown on the y-axis, with bottom measurements at the bottom and upper measurements at the top.

```{r}
#| label: figure-3-stratification-highlight
#| fig-width: 12
#| fig-height: 6

# reorder elevation code so bottom is at the bottom and upper is at the top
do_plot_3 <- do_filtered |> 
  mutate(
    panel_elevation = factor(
      elevation_code,
      levels = c("0", "1", "2")
    )
  )

# set site order for panels
site_levels <- c("East Channel", "Phelps Bridge", "Venoco Bridge")

# create one row per panel for the colored backgrounds
panel_background <- do_plot_3 |> 
  distinct(site_full) |> 
  rename(panel_site = site_full) |> 
  mutate(panel_site = factor(panel_site, levels = site_levels))

# create gray background points by repeating all data across site panels
highlight_background <- do_plot_3 |> 
  
  # cross each observation with each panel site
  crossing(panel_background) |> 
  
  # remove the focal site so only the other sites remain in gray
  filter(site_full != panel_site) |> 
  
  # keep site panel order consistent
  mutate(panel_site = factor(panel_site, levels = site_levels))

# create focal points so each site appears in its own panel
highlight_focal <- do_plot_3 |> 
  mutate(
    panel_site = factor(site_full, levels = site_levels)
  )

# make highlighted stratification plot
figure_3 <- ggplot() +
  
  # blue background identifies each focal site panel
  geom_rect(
    data = panel_background,
    aes(
      xmin = -Inf,
      xmax = Inf,
      ymin = -Inf,
      ymax = Inf,
      fill = panel_site
    ),
    inherit.aes = FALSE,
    alpha = 0.45,
    show.legend = FALSE
  ) +
  
  # gray points show other sites for comparison
  geom_point(
    data = highlight_background,
    aes(
      x = do_mg_l,
      y = panel_elevation
    ),
    color = "gray33",
    alpha = 0.30,
    size = 1.3,
    position = position_jitter(
      width = 0,
      height = 0.12,
      seed = 193
    )
  ) +
  
  # white points show the focal site in each panel
  geom_point(
    data = highlight_focal,
    aes(
      x = do_mg_l,
      y = panel_elevation
    ),
    color = "white",
    alpha = 0.85,
    size = 1.9,
    position = position_jitter(
      width = 0,
      height = 0.12,
      seed = 193
    )
  ) +
  
  # one panel for each site
  facet_wrap(~ panel_site, nrow = 1) +
  
  # use the same site colors as the other figures
  scale_fill_manual(
    values = site_colors
  ) +
  
  # keep the DO axis starting at zero
  scale_x_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  
  # relabel elevation code on the y-axis
  scale_y_discrete(
    labels = c(
      "0" = "Bottom (0)",
      "1" = "Middle (1)",
      "2" = "Upper (2)"
    )
  ) +
  
  # add clear labels
  labs(
    title = "Dissolved Oxygen Stratification by Site",
    x = "Dissolved oxygen (mg/L)",
    y = "Elevation code"
  ) +
  
  # use a simple theme for readability
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold"),
    strip.background = element_rect(fill = "white"),
    panel.border = element_rect(color = "black", fill = NA),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

# show figure
figure_3
```

**Interpretation:** Stratification is most visible at Venoco Bridge, where DO is generally lower at the bottom and higher in the middle and upper elevations.

## Figure 4: Dissolved oxygen through time by elevation code

**Caption:** Figure 4 shows dissolved oxygen measurements through time, separated by elevation code. Points represent individual dissolved oxygen observations, and colors represent sites.

```{r}
#| label: figure-4-do-through-time
#| fig-width: 10
#| fig-height: 8

# reorder elevation code so upper is at the top and bottom is at the bottom
do_plot_4 <- do_filtered |> 
  mutate(
    panel_elevation = factor(
      elevation_code,
      levels = c("2", "1", "0")
    )
  )

# plot dissolved oxygen through time, separated by elevation code and colored by site
figure_4 <- ggplot(
  data = do_plot_4,
  mapping = aes(
    x = date,
    y = do_mg_l,
    color = site_full
  )
) +
  # points show individual DO observations through time
  geom_point(
    alpha = 0.55,
    size = 1.7,
    position = position_jitter(
      width = 2,
      height = 0,
      seed = 193
    )
  ) +
  # panel rows show elevation position
  facet_grid(
    rows = vars(panel_elevation),
    labeller = labeller(
      panel_elevation = c(
        "2" = "Upper (2)",
        "1" = "Middle (1)",
        "0" = "Bottom (0)"
      )
    )
  ) +
  # use the same site colors as the other figures
  scale_color_manual(
    values = site_colors,
    name = "Site"
  ) +
  # keep the DO axis starting at zero
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  # add clear labels
  labs(
    title = "Dissolved Oxygen Through Time by Elevation Code",
    x = "Sampling date",
    y = "Dissolved oxygen (mg/L)"
  ) +
  # use a simple theme for readability
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

# show figure
figure_4
```

**Interpretation:** DO changed through time, but the site pattern was still visible. East Channel was generally higher, Phelps Bridge was generally lower, and Venoco Bridge showed more separation between bottom, middle, and upper measurements.

# 7. Statistical analysis

The statistical analyses follow the class workflow for comparing more than two groups: summarize the data, visualize the pattern, run a Kruskal-Wallis test, use Dunn post-hoc tests when the Kruskal-Wallis result is significant, calculate effect size, and interpret the result. The main response variable is dissolved oxygen. The grouping variables are site and elevation code.

## Question 1: How does dissolved oxygen differ across sites?

This analysis tests whether dissolved oxygen differs across East Channel, Phelps Bridge, and Venoco Bridge.

### Kruskal-Wallis test

```{r}
#| label: stats-site-kruskal

# test whether DO differs across the three main sites
do_site_test <- kruskal.test(
  do_mg_l ~ site_full,
  data = do_filtered
)

# show Kruskal-Wallis test result
do_site_test
```

The Kruskal-Wallis test showed that dissolved oxygen differed significantly across sites, $\chi^2$ = 45.67, df = 2, p < 0.001. This means that DO was not distributed the same way across East Channel, Phelps Bridge, and Venoco Bridge.

### Dunn post-hoc test

```{r}
#| label: stats-site-dunn

# compare site pairs after significant Kruskal-Wallis result
do_site_dunn <- do_filtered |> 
  dunn_test(
    do_mg_l ~ site_full,
    p.adjust.method = "holm"
  )

# show Dunn post-hoc test result
do_site_dunn
```

Because the Kruskal-Wallis test was significant, Dunn post-hoc tests were used to compare site pairs. All three site pairs were significantly different after adjustment: East Channel and Phelps Bridge, East Channel and Venoco Bridge, and Phelps Bridge and Venoco Bridge.

### Effect size

```{r}
#| label: stats-site-effect-size

# calculate effect size for the site comparison
do_site_effect_size <- do_filtered |> 
  kruskal_effsize(
    do_mg_l ~ site_full
  )

# show effect size result
do_site_effect_size
```

The effect size for the site comparison was $\eta^2[H]$ = 0.083, which was interpreted as moderate in the test output. This suggests that site explained a moderate amount of variation in DO ranks in this filtered dataset.

### Interpretation

Dissolved oxygen differed across the three main NCOS sites. East Channel had the highest median DO, Phelps Bridge had the lowest median DO, and Venoco Bridge had the widest spread of values. The Dunn post-hoc test showed that all site pairs were significantly different, so DO patterns were not the same across sites.

## Question 2: How does dissolved oxygen stratification differ across sites?

This analysis tests whether dissolved oxygen differs across elevation-code measurement positions within each site. This addresses stratification because elevation code represents the measurement position in the water column.

### Kruskal-Wallis tests within each site

```{r}
#| label: stats-stratification-within-sites-kruskal

# test DO differences across elevation codes within East Channel
east_channel_elevation_test <- kruskal.test(
  do_mg_l ~ elevation_code,
  data = do_filtered |> 
    filter(site_full == "East Channel")
)

# test DO differences across elevation codes within Phelps Bridge
phelps_bridge_elevation_test <- kruskal.test(
  do_mg_l ~ elevation_code,
  data = do_filtered |> 
    filter(site_full == "Phelps Bridge")
)

# test DO differences across elevation codes within Venoco Bridge
venoco_bridge_elevation_test <- kruskal.test(
  do_mg_l ~ elevation_code,
  data = do_filtered |> 
    filter(site_full == "Venoco Bridge")
)

# show within-site Kruskal-Wallis results
east_channel_elevation_test
phelps_bridge_elevation_test
venoco_bridge_elevation_test
```

The within-site Kruskal-Wallis tests showed no significant DO differences across elevation codes at East Channel or Phelps Bridge. Venoco Bridge did show a significant difference across elevation codes, $\chi^2$ = 28.64, df = 2, p < 0.001. This means that Venoco Bridge had the clearest statistical evidence of DO stratification.

### Dunn post-hoc test for Venoco Bridge

```{r}
#| label: stats-venoco-stratification-dunn

# compare elevation code pairs within Venoco Bridge
venoco_bridge_dunn <- do_filtered |> 
  
  # keep only Venoco Bridge because it had a significant Kruskal-Wallis result
  filter(site_full == "Venoco Bridge") |> 
  
  # compare all elevation-code pairs
  dunn_test(
    do_mg_l ~ elevation_code,
    p.adjust.method = "holm"
  )

# show Venoco Bridge Dunn post-hoc result
venoco_bridge_dunn
```

Because Venoco Bridge had a significant Kruskal-Wallis result, Dunn post-hoc tests were used to compare elevation-code pairs at that site. All three pairs were significantly different after adjustment: bottom and middle, bottom and upper, and middle and upper. This supports the pattern of higher median DO from bottom to upper measurements at Venoco Bridge.

### Effect size for Venoco Bridge

```{r}
#| label: stats-venoco-stratification-effect-size

# calculate effect size for elevation-code differences within Venoco Bridge
venoco_bridge_effect_size <- do_filtered |> 
  
  # keep only Venoco Bridge because it showed significant elevation-code differences
  filter(site_full == "Venoco Bridge") |> 
  
  # calculate Kruskal-Wallis effect size
  kruskal_effsize(
    do_mg_l ~ elevation_code
  )

# show Venoco Bridge effect size result
venoco_bridge_effect_size
```

The effect size for the Venoco Bridge elevation-code comparison was $\eta^2[H]$ = 0.105, which was interpreted as moderate in the test output. This suggests that elevation code explained a moderate amount of variation in DO ranks at Venoco Bridge.

### Interpretation

Statistical evidence for DO stratification was clearest at Venoco Bridge. East Channel and Phelps Bridge did not show significant DO differences across elevation codes, while Venoco Bridge did. At Venoco Bridge, median DO was lower at the bottom and higher at the middle and upper positions.

# 8. Overall interpretation

Overall, dissolved oxygen differed across the three main NCOS sites, and statistical evidence for stratification was clearest at Venoco Bridge. East Channel had the highest median DO, Phelps Bridge had the lowest median DO, and Venoco Bridge showed the clearest increase from bottom to upper measurements. These results suggest that DO conditions and stratification patterns were not the same across the three sites.